Towards automated provenance collection for experimental runs of agent-based models\thanks{This work was supported by the Scottish Government Rural and Environment Science and Analytical Services Division (project reference JHI-C5-1)}

### 2023-04-19 paper/paper.tex
Updates from Overleaf

Replication of social simulation results has been highlighted as a significant
issue for the ABM community for a number of years (e.g. \cite{edmonds2003replication}).
The paper that forms the basis of this work (\cite{polhill2017lessons}) shows that
the analysis of the outputs of the model can potentially be just as complex a
process as developing and running the model itself. Analysis of outputs is no less difficult
to replicate unless adequate records are kept. The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook. There are also 
many scientific workflow tools, such as Snakemake
\cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
\cite{ludascher2006scientific} and Taverna \cite{hull2006taverna}. However, since
these are workflow tools, the focus is on automation and
repeatability rather than provenance, which, if it is included, is as an afterthought.

@@

The output analysis replication in this paper concerns earlier work with
FEARLUS-SPOMM, which is a coupled agent-based model of agricultural decision-
making and species stochastic patch occupancy metacommunity model that has been
used to explore incentivisation strategies to improve biodiversity
\cite{polhill_nonlinearities_2013,gimona2011exploring}. Belonging to the
‘typification’ class of social simulations  \cite{boero2005does}, this work
involved the analysis of the outputs of approximately 20,000 runs of the model
using a number of techniques aimed at demonstrating nonlinearities in the
relationship between incentivisation and biodiversity outcome.  Recording
provenance metadata on the process used to analyse and visualise the outputs is challenging, and
currently there are no codified standards as to how this should be done for
agent-based models. For FEARLUS-SPOMM, the analysis and visualisation methods used drew heavily on statistical
techniques available as R packages.
Although R allows transcripts of interactive terminal sessions to be saved, the
work involved great deal of exploration of alternative analyses and visualisations, not all of
which were reported in the manuscript as finally accepted. Such transcripts are therefore not the
best way to record the means by which a model's outputs were analysed, and hence the
strategy used was to save each analysis or visualisation in a(n R) script.
Since the output from the (Swarm) model software used a mixture of text formats, some Perl scripts were also written to
process that output into CSV format for easy import into R. When the MIRACLE
project \cite{parker2019final} provided a context in which the replication of
that analysis was necessary, an opportunity was created to test the viability
of the strategy of relying on scripts to record provenance.

`Multifinality' (the same initial conditions and parameter settings having qualitatively different results) in agent-based models means we need to ensure that
reported results are not just down to a matter of chance. This is one of many reasons (e.g. in empirical contexts especially, calibration, validation and sensitivity analysis) why experiments with ABMs involve large numbers of runs. Most ABM experiments therefore resort to some kind of automation, including of the kind provided by workflow tools mentioned above, but also using built-in features of ABM software (e.g. in the case of NetLogo, BehaviorSpace). We refer to this
as the automatic and replicability problem. In addition to this there is the
problem of recording which data is being used, which data is being produced,
how it was gathered or created and how the resultant data was processed. Given
20,000 experiments the amount of such data is not inconsiderable. Moreover this
data needs to be in a form that is analysable and visualisable.

@@

\begin{itemize} \item {\color{green} \ttvar{create_database.py}} - creates a
        database idempotently.  \item {\color{green} \ttvar{exists.py}} - checks
        if a particular row in a table exists.  \item {\color{green}
            \ttvar{get_value.py}} - gets any specified single value from a
            table given the primary key.  \item {\color{green}
\ttvar{get_values.py}} \item {\color{green} \ttvar{next_study.py}} \item
{\color{green} \ttvar{update.py}} - idempotently updates a particular row in a
table.  \end{itemize}

@@

So each of the Bash commands listed in table \ref{tab:commmands} composes over
the above python commands to populate the SSREPI database in a consistent and
logical manner.

>>>>

Replication of social simulation results has been highlighted as a significant
issue for the agent-based modelling (ABM) community for a number of years (e.g. \cite{edmonds2003replication}).
The paper that forms the basis of this work (\cite{polhill2017lessons}) shows that
the analysis of the outputs of the model can potentially be just as complex a
process as developing and running the model itself. Analysis of outputs is no less difficult
to replicate unless adequate records are kept. The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook. There are also 
many scientific workflow tools, such as Snakemake
\cite{koster2012snakemake}, NextFlow \cite{di2017nextflow}, Kepler
\cite{ludascher2006scientific} and Taverna \cite{hull2006taverna}. However, since
these are workflow tools, the focus is on automation and
repeatability rather than provenance, which, if it is included, is as an afterthought.

@@

The output analysis replication in this paper concerns earlier work with
FEARLUS-SPOMM, which is a coupled ABM of agricultural decision-
making and species stochastic patch occupancy metacommunity model that has been
used to explore incentivisation strategies to improve biodiversity
\cite{polhill_nonlinearities_2013,gimona2011exploring}. Belonging to the
‘typification’ class of social simulations  \cite{boero2005does}, this work
involved the analysis of the outputs of approximately 20,000 runs of the model
using a number of techniques aimed at demonstrating nonlinearities in the
relationship between incentivisation and biodiversity outcome.  Recording
provenance metadata on the process used to analyse and visualise the outputs is challenging, and
currently there are no codified standards as to how this should be done for
ABMs. For FEARLUS-SPOMM, the analysis and visualisation methods used drew heavily on statistical
techniques available as R packages.
Although R allows transcripts of interactive terminal sessions to be saved, the
work involved great deal of exploration of alternative analyses and visualisations, not all of
which were reported in the manuscript as finally accepted. Such transcripts are therefore not the
best way to record the means by which a model's outputs were analysed, and hence the
strategy used was to save each analysis or visualisation in a(n R) script.
Since the output from the (Swarm) model software used a mixture of text formats, some Perl scripts were also written to
process that output into CSV format for easy import into R. When the MIRACLE
project \cite{parker2019final} provided a context in which the replication of
that analysis was necessary, an opportunity was created to test the viability
of the strategy of relying on scripts to record provenance.

`Multifinality' (the same initial conditions and parameter settings having qualitatively different results) in ABMs means we need to ensure that
reported results are not just down to a matter of chance. This is one of many reasons (e.g. in empirical contexts especially, calibration, validation and sensitivity analysis) why experiments with ABMs involve large numbers of runs. The tedium (and in larger-scale cases, infeasibility) of conducting each run by hand means most ABM experiments resort to some kind of automation, including of the kind provided by workflow tools mentioned above, but also using built-in features of ABM software (e.g. in the case of NetLogo, BehaviorSpace). This is fine if all we
want to do is repeat the same process, but what if we are interested in why a particular instance of that process led to an unexpected outcome? Re-executing the workflow will not necessarily generate \textit{exactly} the same result unless we have a record of everything needed to do that (including seeds for pseudo-random number generators). We refer to this
as the `automation and replication' problem.

%%% GP: Got here as at 2023-04-19T10:59.

@@

\begin{itemize} \item {\color{green} \ttvar{create_database.py}} - creates a
        database idempotently.  \item {\color{green} \ttvar{exists.py}} - checks
        if a particular row in a table exists.  \item {\color{green}
            \ttvar{get_value.py}} - gets any specified single value from a
            table given the primary key.  \item {\color{green}
\ttvar{get_values.py}}  - gets one or more rows given the search values. \item {\color{green}\ttvar{next_study.py}} - gets the next available and unique study number. \item
{\color{green} \ttvar{update.py}} - idempotently updates a particular row in a
table.  \end{itemize}

@@

So each of the Bash commands listed in table \ref{tab:commands} composes over
the above python commands to populate the SSREPI database in a consistent and
logical manner.
